---
title: "Final_Presentation_Gabriel"
author: "Gabriel"
format: pdf
editor: visual
---

```{r}
library(tidyverse) # For data wrangling
library(brms) # For stats
library(ggeffects) # for plotting model predictions
# Note: I needed to also install the `insight` and `see` packages to get `modelbased` to install properly; try that if you get similar error messages 
#install.packages('insight')
library(insight)
#install.packages('see')
library(see)
#install.packages('modelbased') # if you need to install this package
library(modelbased) # for plotting model predictions. supports the link scale (ggeffects does not)
#install.packages('faraway') # if you need to install this package
library(faraway) # For data on galapagos species richess
library(ggplot2)
library(here)
#install.packages("dagitty")   # only the first time
library(dagitty)
```

## Data Wrangling

```{r}
#This chunk is intended to combine the no3 and climind df
# 1. Read the CSV files into R
kelp_data <- read.csv("rawdata/kelp_no3_waves_quarterly_long.csv")
clim_data <- read.csv("rawdata/climind_quarterly.csv")

# 2. Combine the data frames
combined_sitespecific_climind <- left_join(kelp_data, clim_data, by = c("year", "quarter"))

# View the new dataframe
head(combined_sitespecific_climind)
```

```{r}
# Save the combined data frame as a new CSV file
write.csv(combined_sitespecific_climind, file = "newdf/combined_kelp_climind_data.csv", row.names = FALSE)
```

```{r}
#this chunk is to create a new df with temp included
# 1. Read the temperature data
temp_data <- read.csv("rawdata/SBH_temperature_20241231.csv")

# 2. Convert daily data to quarterly averages
temp_quarterly <- temp_data %>%
  # Create a "quarter" column based on the MONTH (e.g., Jan-Mar = 1, Apr-Jun = 2, etc.)
  mutate(quarter = ceiling(MONTH / 3)) %>%
  # Rename YEAR to lowercase "year" so it matches your other datasets
  rename(year = YEAR) %>%
  # Group by the year and our newly created quarter
  group_by(year, quarter) %>%
  # Calculate the average surface temperature for that quarter (ignoring missing data)
  summarize(mean_surf_temp = mean(SURF_TEMP_C, na.rm = TRUE), .groups = "drop")

# 3. Join the new quarterly temperature data to your combined dataframe
combined_sitespecific_climind_temp <- left_join(combined_sitespecific_climind, temp_quarterly, by = c("year", "quarter"))

# View the results to make sure 'mean_surf_temp' was added
head(combined_sitespecific_climind_temp)

#save df as file
write.csv(combined_sitespecific_climind_temp, file = "newdf/combined_sitespecific_climind_temp.csv", row.names = FALSE)
```

## Data Searching

### NO3 Vs. Kelp Biomass

```{r}
combined_data = read.csv(here("newdf/combined_sitespecific_climind_temp.csv"))


p.nitrateVsbiomass = ggplot(combined_data, aes(x = no3, y = kelp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ no3, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
  labs(
    title = "Relationship Between Nitrate Concentrations and Kelp Biomass",
    x = "Nitrate Concentration (µmol/L)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()


p.nitrateVsbiomass
```

### Wave Height Vs. Kelp Biomass

```{r}
#Creates a scatter plot to show the relationship between waves and biomass

p.wavesVsbiomass = ggplot(combined_data, aes(x = waves, y = kelp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ waves, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
  labs(
    title = "Relationship Between Wave Height and Kelp Biomass",
    x = "Wave Height (M)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()


p.wavesVsbiomass
```

### Mean Surface Temperature Vs. Kelp Biomass

```{r}
p.tempVsbiomass = ggplot(combined_data, aes(x = mean_surf_temp, y = kelp)) +
  geom_point() +
  geom_smooth(method = "lm") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ mean_surf_temp, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
  labs(
    title = "Relationship Between Mean Surface Temperature and Kelp Biomass",
    x = "Mean Surface Temperature (°C)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()


p.tempVsbiomass
```

### DAG

```{r}
kelp_dag <- dagitty('dag {
bb="-1.707,-4.687,2.5,4.206"
Eutrophication [latent,pos="0.778,0.628"]
Grazing [latent,pos="-0.694,1.288"]
Kelp_Biomass [outcome,pos="0.074,1.253"]
MEI [pos="-0.768,-0.889"]
NPGO [pos="0.392,-1.792"]
Nitrate [pos="0.079,-0.136"]
PDO [pos="-0.597,-1.734"]
Temperature [pos="0.812,-0.727"]
Wave_Height [pos="-0.738,0.535"]

Eutrophication -> Nitrate
Grazing -> Kelp_Biomass
MEI -> Nitrate
MEI -> Temperature
MEI -> Wave_Height
NPGO -> Nitrate
NPGO -> Temperature
Nitrate -> Kelp_Biomass
PDO -> Nitrate
PDO -> Temperature
Temperature -> Kelp_Biomass
Temperature -> Nitrate
Wave_Height -> Kelp_Biomass
Wave_Height -> Nitrate
}')

plot(kelp_dag)
impliedConditionalIndependencies(kelp_dag)
```

#### Multiple Regression

##### What variables are direct causes of nitrate concentrations according to our DAG? (Adjustment direct effect)

```{r}
#Including NPGO + PDO + MEI + mean_surf_temp
#estimating the direct effect of NPGO, PDO, Temperature, and MEI on Nitrate:
m.nitrate.MEI.NPGO.PDO.Temp <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      no3 ~ NPGO + PDO + MEI + mean_surf_temp,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.nitrate.MEI.NPGO.PDO.Temp")


summary(m.nitrate.MEI.NPGO.PDO.Temp)
plot(m.nitrate.MEI.NPGO.PDO.Temp)
```

#### What variables are direct causes of mean surface temperature according to our DAG? (Adjustment direct effect)

```{r}
#Including MEI 
#estimating the direct effect of MEI on surface temp:
m.temp.MEI <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      mean_surf_temp ~ MEI,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.temp.MEI")


summary(m.temp.MEI)
plot(m.temp.MEI)
```

```{r}
#Including PDO
#estimating the direct effect of PDO on surface temp:
m.temp.PDO <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      mean_surf_temp ~ PDO,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.temp.PDO")


summary(m.temp.PDO)
plot(m.temp.PDO)
```

```{r}
#Including NPGO
#estimating the direct effect of NPGO on surface temp:
m.temp.NPGO <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      mean_surf_temp ~ NPGO,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.temp.NPGO")


summary(m.temp.NPGO)
plot(m.temp.NPGO)
```

##### What variables are direct causes of kelp biomass according to our DAG? (Adjustment direct effect)

```{r}
#Including temp + nitrate + waves
#finding the direct effect of mean surface temp, nitrate concentration, and wave height on Biomass:
m.biomass.temp.nitrate.waves <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      kelp ~ mean_surf_temp + no3 + waves,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.biomass.temp.nitrate.waves")

summary(m.biomass.temp.nitrate.waves)
plot(m.biomass.temp.nitrate.waves)
```

##### What are the total effects of NPGO, MEI, and PDO on Kelp biomass

```{r}
#Total effect of NPGO on kelp biomass
m.kelp.NPGO.total <- 
  brm(
    data = combined_sitespecific_climind_temp, #gives data
    family = gaussian, #sets family as gaussian distribution
    kelp ~ NPGO, #Sets parameters based on dag
    iter = 2000, warmup = 1000, chains = 4, cores = 4, 
    seed = 4,
    file = "outputs/m.kelp.NPGO.total"
  )

summary(m.kelp.NPGO.total)
plot(m.kelp.NPGO.total)
```

```{r}
#Total effect of PDO on kelp biomass
m.kelp.PDO.total <- 
  brm(
    data = combined_sitespecific_climind_temp, #gives data
    family = gaussian, #sets family as gaussian distribution
    kelp ~ PDO, #Sets parameters based on dag
    iter = 2000, warmup = 1000, chains = 4, cores = 4, 
    seed = 4,
    file = "outputs/m.kelp.PDO.total"
  )

summary(m.kelp.PDO.total)
plot(m.kelp.PDO.total)
```

```{r}
#Total effect of MEI on kelp biomass
m.kelp.MEI.total <- 
  brm(
    data = combined_sitespecific_climind_temp, #gives data
    family = gaussian, #sets family as gaussian distribution
    kelp ~ MEI, #Sets parameters based on dag
    iter = 2000, warmup = 1000, chains = 4, cores = 4, 
    seed = 4,
    file = "outputs/m.kelp.MEI.total"
  )

summary(m.kelp.MEI.total)
plot(m.kelp.MEI.total)
```
