---
title: "Final_Presentation_Gabriel"
author: "Gabriel"
format: pdf
editor: visual
---

```{r}
#install.packages('tidyverse')
library(tidyverse) # For data wrangling
#install.packages('brms')
library(brms) # For stats
#install.packages('ggeffects')
library(ggeffects) # for plotting model predictions
# Note: I needed to also install the `insight` and `see` packages to get `modelbased` to install properly; try that if you get similar error messages 
#install.packages('insight')
library(insight)
#install.packages('see')
library(see)
#install.packages('modelbased') # if you need to install this package
library(modelbased) # for plotting model predictions. supports the link scale (ggeffects does not)
#install.packages('faraway') # if you need to install this package
library(faraway) # For data on galapagos species richess
#install.packages('ggplot2')
library(ggplot2)
#install.packages('here')
library(here)
#install.packages('dagitty')   # only the first time
library(dagitty)
```

## Data Wrangling

First we are going to combine the data into one data frame. This will make using the data later on easier as we will only have to refer to one data frame when creating plots and models. Left join will be great for this as it will ensure that all data ends up in the proper location. There were some issues with the way the data moved when done at once, so it has been broken down into several steps. Luckily, we will not have to pivot any data as it is already in the proper format.

```{r}
#This chunk is intended to combine the no3, climind df, and temp data frame
#Read the CSV files into R
kelp_data <- read.csv("rawdata/kelp_no3_waves_quarterly_long.csv")
clim_data <- read.csv("rawdata/climind_quarterly.csv")

#Combine the data frames
combined_sitespecific_climind <- left_join(kelp_data, clim_data, by = c("year", "quarter"))

#View the new dataframe
head(combined_sitespecific_climind)

#Save the combined data frame as a new CSV file
write.csv(combined_sitespecific_climind, file = "newdf/combined_kelp_climind_data.csv", row.names = FALSE)

#Read the temperature data
temp_data <- read.csv("rawdata/SBH_temperature_20241231.csv")

#Convert daily data to quarterly averages
temp_quarterly <- temp_data %>%
  # Create a "quarter" column based on the MONTH (e.g., Jan-Mar = 1, Apr-Jun = 2, etc.)
  mutate(quarter = ceiling(MONTH / 3)) %>%
  # Rename YEAR to lowercase "year" so it matches your other datasets
  rename(year = YEAR) %>%
  # Group by the year and our newly created quarter
  group_by(year, quarter) %>%
  # Calculate the average surface temperature for that quarter (ignoring missing data)
  summarize(mean_surf_temp = mean(SURF_TEMP_C, na.rm = TRUE), .groups = "drop")

#Join the new quarterly temperature data to your combined dataframe
combined_sitespecific_climind_temp <- left_join(combined_sitespecific_climind, temp_quarterly, by = c("year", "quarter"))

#View the results
head(combined_sitespecific_climind_temp)

#save df as file
write.csv(combined_sitespecific_climind_temp, file = "newdf/combined_sitespecific_climind_temp.csv", row.names = FALSE)
combined_data = read.csv(here("newdf/combined_sitespecific_climind_temp.csv"))
```

## Data Searching

We are going to plot several different relationships to see how they may correlate. This alone will not give us enough information to draw conclusions, but it will help us learn where to start looking.

Each relationship will be demonstrated on two separate plots. In the first plot the data will be modeled by both geom_smooth model "lm" and geom_smooth model "gam" ("gam" is used instead of "loess" as loss would not properly dsiplay. We believe that this is due to the sheer number of samples). When looking at the plot we can see how well the "gam" model fits with the "lm" model. If they are close together then the data is likely linear, if they deviate strongly then it is likely non-linear. The second plot for each relationship will be on a log scale. This is done to see if the relationship is logarithmic. If the slope of the data appears to become linear in these models than we know that the relationship is logarithmic.

### NO3 vs Kelp Biomass

Here we run the relationships for nitrate Vs Kelp biomass.

```{r}
#Ploting no3 vs biomass to find if the data is linear or nonlinear
p.nitrateVsbiomass.linearitycheck = ggplot(combined_data, aes(x = no3, y = kelp)) +
  stat_summary_bin(
    bins = 60,
    fun = mean,
    geom = "point"
  )+
  labs(
    title = "Relationship Between Nitrate Concentrations and Kelp Biomass",
    x = "Nitrate Concentration (µmol/L)",
    y = "Kelp Biomass (Kg)"
  )+
  theme_classic()+
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ no3, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2)+
  geom_smooth(method = "gam", formula = y ~ s(x) , se = TRUE, linetype = "dashed", color = "green", linewidth = 1)

ggsave(
  filename = here("outputs", "p.nitrateVsbiomass.linearitycheck.png"),
  plot = p.nitrateVsbiomass.linearitycheck
)

p.nitrateVsbiomass.linearitycheck 
```

```{r}
#Ploting nitrate vs biomass against a log scale
p.nitrateVsbiomass.logscale = ggplot(combined_data, aes(x = no3, y = kelp)) +
  stat_summary_bin(
    bins = 60,
    fun = mean,
    geom = "point"
  ) +
  scale_x_log10() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  theme_classic() +
  labs(
    title = "Kelp Biomass vs Nitrate Concentration",
    x = "Nitrate Concentration (log10 scale)",
    y = "Kelp Biomass (kg)"
  )

ggsave(
  filename = here("outputs", "p.nitrateVsbiomass.logscale.png"),
  plot = p.nitrateVsbiomass.logscale
)

p.nitrateVsbiomass.logscale
```

We can see that the biomass certainly changes as the nitrate concentration changes. On the first plot we see that the data takes more of an M shape. On the second plot we do see that the data begins to make a more linear path. For this reason the relationship between nitrate and Vs kelp biomass is approximately logarithmic.

### Wave Height Vs. Kelp Biomass

Here we are creating plots to model wave height Vs. biomass

```{r}
#Creates a scatter plot to show the relationship between waves and biomass

p.wavesVsbiomass = ggplot(combined_data, aes(x = waves, y = kelp)) +
  stat_summary_bin(
    bins = 60,
    fun = mean,
    geom = "point"
  )+
  geom_smooth(method = "lm", color = "red") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ waves, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
    geom_smooth(method = "gam", formula = y ~ s(x) , se = TRUE, linetype = "dashed", color = "green", linewidth = 1)+
  labs(
    title = "Relationship Between Wave Height and Kelp Biomass",
    x = "Wave Height (M)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()

ggsave(
  filename = here("outputs", "p.wavesVsbiomass.png"),
  plot = p.wavesVsbiomass
)

p.wavesVsbiomass
```

```{r}
#Creates a scatter plot to show the relationship between waves and biomass with a log scale

p.wavesVsbiomass.logarithmic = ggplot(combined_data, aes(x = waves, y = kelp)) +
  stat_summary_bin(
    bins = 60,
    fun = mean,
    geom = "point"
  )+
  geom_smooth(method = "lm", color = "red") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ waves, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
  scale_x_log10() +
    geom_smooth(method = "gam", formula = y ~ s(x) , se = TRUE, linetype = "dashed", color = "green", linewidth = 1)+
  labs(
    title = "Relationship Between Wave Height and Kelp Biomass",
    x = "Wave Height (M)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()

ggsave(
  filename = here("outputs", "p.wavesVsbiomass.logarithmic.png"),
  plot = p.wavesVsbiomass.logarithmic
)

p.wavesVsbiomass.logarithmic
```

### Mean Surface Temperature Vs. Kelp Biomass

```{r}
p.tempVsbiomass = ggplot(combined_data, aes(x = mean_surf_temp, y = kelp)) +
stat_summary_bin(
    bins = 30,
    fun = mean,
    geom = "point"
  )+  
  geom_smooth(method = "lm") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ mean_surf_temp, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
  labs(
    title = "Relationship Between Mean Surface Temperature and Kelp Biomass",
    x = "Mean Surface Temperature (°C)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()

ggsave(
  filename = here("outputs", "p.tempVsbiomass.png"),
  plot = p.tempVsbiomass
)

p.tempVsbiomass
```

```{r}
#Ploting nitrate vs biomass against a log scale
#Not very informative because there is such a narrow temp range
p.tempVsbiomass.log = ggplot(combined_data, aes(x = mean_surf_temp, y = kelp)) +
stat_summary_bin(
    bins = 30,
    fun = mean,
    geom = "point"
  )+  
  geom_smooth(method = "lm") +
  annotate(
    "text",
    x = Inf, y = Inf,
    label = paste("Slope =", round(coef(lm(kelp ~ mean_surf_temp, data = combined_data))[2], 3)),
    hjust = 1.1, vjust = 1.5
  )+
    scale_x_log10() +
  labs(
    title = "Relationship Between Mean Surface Temperature and Kelp Biomass",
    x = "Mean Surface Temperature (°C)",
    y = "Kelp Biomass (Kg)"
  ) +
  theme_classic()

ggsave(
  filename = here("outputs", "p.tempVsbiomass.log.png"),
  plot = p.tempVsbiomass.log
)

p.tempVsbiomass.log
```

### DAG

```{r}
kelp_dag <- dagitty('dag {
bb="-1.707,-4.687,2.5,4.206"
Eutrophication [latent,pos="0.778,0.628"]
Grazing [latent,pos="-0.694,1.288"]
Kelp_Biomass [outcome,pos="0.074,1.253"]
MEI [pos="-0.768,-0.889"]
NPGO [pos="0.392,-1.792"]
Nitrate [pos="0.079,-0.136"]
PDO [pos="-0.597,-1.734"]
Temperature [pos="0.812,-0.727"]
Wave_Height [pos="-0.738,0.535"]

Eutrophication -> Nitrate
Grazing -> Kelp_Biomass
MEI -> Nitrate
MEI -> Temperature
MEI -> Wave_Height
NPGO -> Nitrate
NPGO -> Temperature
Nitrate -> Kelp_Biomass
PDO -> Nitrate
PDO -> Temperature
Temperature -> Kelp_Biomass
Temperature -> Nitrate
Wave_Height -> Kelp_Biomass
Wave_Height -> Nitrate
}')

png(here("outputs", "kelp_dag.png"), width = 1600, height = 1200, res = 300)
plot(kelp_dag)
dev.off()

plot(kelp_dag)
impliedConditionalIndependencies(kelp_dag)
```

#### Multiple Regression

##### What variables are direct causes of nitrate concentrations according to our DAG? (Adjustment direct effect)

```{r}
#Including NPGO + PDO + MEI + mean_surf_temp
#estimating the direct effect of NPGO, PDO, Temperature, and MEI on Nitrate:
m.nitrate.MEI.NPGO.PDO.Temp <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      no3 ~ NPGO + PDO + MEI + mean_surf_temp,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.nitrate.MEI.NPGO.PDO.Temp")


summary(m.nitrate.MEI.NPGO.PDO.Temp)
plot(m.nitrate.MEI.NPGO.PDO.Temp)
```

#### What variables are direct causes of mean surface temperature according to our DAG? (Adjustment direct effect)

```{r}
#Including MEI 
#estimating the direct effect of MEI on surface temp:
m.temp.MEI <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      mean_surf_temp ~ MEI,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.temp.MEI")


summary(m.temp.MEI)
plot(m.temp.MEI)
```

```{r}
#Including PDO
#estimating the direct effect of PDO on surface temp:
m.temp.PDO <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      mean_surf_temp ~ PDO,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.temp.PDO")


summary(m.temp.PDO)
plot(m.temp.PDO)
```

```{r}
#Including NPGO
#estimating the direct effect of NPGO on surface temp:
m.temp.NPGO <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      mean_surf_temp ~ NPGO,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.temp.NPGO")


summary(m.temp.NPGO)
plot(m.temp.NPGO)
```

##### What variables are direct causes of kelp biomass according to our DAG? (Adjustment direct effect)

```{r}
#Including temp + nitrate + waves
#finding the direct effect of mean surface temp, nitrate concentration, and wave height on Biomass:
m.biomass.temp.nitrate.waves <- 
  brm(data = combined_sitespecific_climind_temp, # Give the model the combined data
      # Choose a gaussian (normal) distribution
      family = gaussian,
      # Specify the model here. 
      kelp ~ mean_surf_temp + no3 + waves,
      # Here's where you specify parameters for executing the Markov chains
      # We're using similar to the defaults, except we set cores to 4 so the analysis runs faster than the default of 1
      iter = 2000, warmup = 1000, chains = 4, cores = 4,
      # Setting the "seed" determines which random numbers will get sampled.
      # In this case, it makes the randomness of the Markov chain runs reproducible 
      # (so that both of us get the exact same results when running the model)
      seed = 4,
      # Save the fitted model object as output - helpful for reloading in the output later
      file = "outputs/m.biomass.temp.nitrate.waves")

summary(m.biomass.temp.nitrate.waves)
plot(m.biomass.temp.nitrate.waves)
```

##### What are the total effects of NPGO, MEI, and PDO on Kelp biomass

```{r}
#Total effect of NPGO on kelp biomass
m.kelp.NPGO.total <- 
  brm(
    data = combined_sitespecific_climind_temp, #gives data
    family = gaussian, #sets family as gaussian distribution
    kelp ~ NPGO, #Sets parameters based on dag
    iter = 2000, warmup = 1000, chains = 4, cores = 4, 
    seed = 4,
    file = "outputs/m.kelp.NPGO.total"
  )

summary(m.kelp.NPGO.total)
plot(m.kelp.NPGO.total)
```

```{r}
#Total effect of PDO on kelp biomass
m.kelp.PDO.total <- 
  brm(
    data = combined_sitespecific_climind_temp, #gives data
    family = gaussian, #sets family as gaussian distribution
    kelp ~ PDO, #Sets parameters based on dag
    iter = 2000, warmup = 1000, chains = 4, cores = 4, 
    seed = 4,
    file = "outputs/m.kelp.PDO.total"
  )

summary(m.kelp.PDO.total)
plot(m.kelp.PDO.total)
```

```{r}
#Total effect of MEI on kelp biomass
m.kelp.MEI.total <- 
  brm(
    data = combined_sitespecific_climind_temp, #gives data
    family = gaussian, #sets family as gaussian distribution
    kelp ~ MEI, #Sets parameters based on dag
    iter = 2000, warmup = 1000, chains = 4, cores = 4, 
    seed = 4,
    file = "outputs/m.kelp.MEI.total"
  )

summary(m.kelp.MEI.total)
plot(m.kelp.MEI.total)
```
